SSA based global code motion of identical computations

### 2017-04-17 acmart-pacmpl-template.tex
macro fixes

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \SESE {SESE}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \LLVMTestSuite {\LLVM{} test-suite}
\def \SPEC {SPEC Cpu2006}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We present a global code motion compiler optimization which schedules
identical computations across the program so as to save code size.  Not only
this code motion saves code size, it exposes redundancies in some cases, it
exposes more instruction level parallelism in the basic block when the
computations are hoisted, and it enables other passes like loop invariant
motion to remove more redundancies. The cost model to drive the code motion is
based on live range analysis on SSA representation such that the (virtual)
register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
of 2.5\% savings in code size in llvm testsuite, although the code size also
increases in many cases because it enables more inlining. We have also seen
imrpovements of approximately 5\% in a couple of SPEC 2006 benchmarks viz. gcc
and mcf, moreover, register spills reduced for almost all the SPEC 2006
bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
sense that it considers all identical computations in a function as potential
candidates. We make an extra effort to hoist candidates by partitioning the
potential candidates in a way to enable partial hoisting in case common
hoisting points for all the candidates cannot be found. We also formalize how
register pressure will reduce as a result of code-motion and why sorting the
list of potential candidates w.r.t. their depth first numbers helps hoist more
candidates with less compile time overhead.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Optimizng Compilers, GCC, LLVM, Code Generation, Global Scheduling}  %% \keywords is optional

@@

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in LLVM trunk,
however, a general implementation of code-motion of congruent (but not
redundant) expressions is still missing from GCC and LLVM trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches,
\item a cost model to reduce interference and hence, reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  SSA based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

Shobaki \cite{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which is not feasible for industrial
compilers like gcc and LLVM.

@@

\subsection{Liveness using DJ Graph}
DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-Graph
and Merge Sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic block. The underlying
simplicity of liveness computation is due to SSA form, where the the values only
flow (from def to use) either through dominator edges or the join edges (where
we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge sets for each basic block i.e., a set of all basic blocks
where the values can flow from a particular basic block. We have implemented
merge-set computation from DJ-graph as illustrated in \cite{das2005}. A simplified
version of merge set that we implemented is presented here:

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by GCC's code-hoisting but that would be a lot
of data structures to maintain at each basic block level. For GCC it makes sense
because their code-hoisting is integrated with GVN-PRE which already has those
data structures available.

@@

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the MemorySSA of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and that there are no aliasing
loads or store along the path from the current position to the hoisting point.

@@

Moreover, LLVM has a `getelementptr' instruction which computes the address
where a load or a store would happen. It is a scalar computation and gets
hoisted frequently even if the corresponding loads/stores would not get hoisted.
In order to reduce register pressure while hoisting loads, we have restricted
hoisting of address computations away from their corresponding loads and stores
when the loads and stores cannot be moved.  This restriction is only to mitigate
the limitations of LLVM's register allocator and may be lifted in the future,
when the register allocation rematerialization pass has been improved to catch
these regressions.

@@

\begin{figure}
\centering
%rankdir=LR;
\digraph[scale=0.35]{abc}{
  B0 -> B1
  B0 -> B2
  B1 -> B3
  B1 -> B4
  B2 -> B5
  B2 -> B6
}
\caption{CFG to illustrate partitioning}
\label{fig:dfsin}
\end{figure}

@@

\subsection{Code generation}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed, and the \SSA{} form is restored by updating the
intermediate representation (IR) to reflect the changes. At the same time
MemorySSA is also updated to get up-to-date information about memory references.

@@

\section{SSA based global code motion of identical computations}
\label{sec:implementation-details}
% Move this before legality and profitability??
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna
In this section we will present an overview of the algorithm we implemented in
\LLVM{}.  The amount of hoisting depends on whether we collect \GVN{} of
instructions before finding candidates (optimistic) or, on-demand
(pessimistic). It also depends on the generality of the \GVN{} algorithm as
mentioned earlier in Section~\ref{subsec:finding-candidates}. We have
implemented a optimistic global code motion of congruent instructions which uses
the liveness analysis as illustrated in Das \cite{das2012} and ranking
expressions explained in Rosen \cite{rosen1988global}.

@@

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to LLVM IR. Since a basic block in LLVM IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot not guarantee progress is marked
as a barrier. In the absense of context (as in our current implementation), some
instructions which might be safe are still classified a barrier e.g., volatile
loads/stores, calls with missing attributes.

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but LLVM's GVN infrastucture does not compute
equivalence classes in a effective way. We found it simpler to just recompute
the value numbers and start finding hoistable candidates again.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ Graph and MergeSets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis expensive but only performed
on-demand for hoistable candidates so it does not affect the overall compile
time by much. Other analyses like Alias Analysis, Memory SSA and Dominator Tree
are already available in the LLVM pass pipeline. Although we recompute GVN, and
Barriers for each iteration of the code-motion, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite code-motion by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

@@

While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined. Various code-size metrics are shown in
Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
instructions were sunk.

>>>>

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \SPEC {SPEC Cpu2006}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We present a global code motion compiler optimization which schedules
identical computations across the program so as to save code size.  Not only
this code motion saves code size, it exposes redundancies in some cases, it
exposes more instruction level parallelism in the basic block when the
computations are hoisted, and it enables other passes like loop invariant
motion to remove more redundancies. The cost model to drive the code motion is
based on live range analysis on \SSA{} representation such that the (virtual)
register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
of 2.5\% savings in code size on \SPEC{}, although the code size also
increases in many cases because it enables more inlining. We have also seen
execution time improvements of approximately 5\% in a couple of \SPEC{} benchmarks viz. \GCC{}
and mcf, moreover, register spills reduced for almost all the \SPEC{}
bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
sense that it considers all identical computations in a function as potential
candidates. We make an extra effort to hoist candidates by partitioning the
potential candidates in a way to enable partial hoisting in case common
hoisting points for all the candidates cannot be found. We also formalize how
register pressure will reduce as a result of code-motion and why sorting the
list of potential candidates w.r.t. their depth first numbers helps hoist more
candidates with less compile time overhead.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Optimizng Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional

@@

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
however, a general implementation of code-motion of congruent (but not
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches,
\item a cost model to reduce interference and hence, reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

Shobaki \cite{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which is not feasible for industrial
compilers like \GCC{} and \LLVM{}.

@@

\subsection{Liveness using DJ Graph}
DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-Graph
and Merge Sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values only
flow (from def to use) either through dominator edges or the join edges (where
we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge sets for each basic block i.e., a set of all basic blocks
where the values can flow from a particular basic block. We have implemented
merge-set computation from DJ-graph as illustrated in \cite{das2005}. A simplified
version of merge set that we implemented is presented here:

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
of data structures to maintain at each basic block level. For \GCC{} it makes sense
because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
data structures available.

@@

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and that there are no aliasing
loads or store along the path from the current position to the hoisting point.

@@

Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
where a load or a store would happen. It is a scalar computation and gets
hoisted frequently even if the corresponding loads/stores would not get hoisted.
In order to reduce register pressure while hoisting loads, we have restricted
hoisting of address computations away from their corresponding loads and stores
when the loads and stores cannot be moved.  This restriction is only to mitigate
the limitations of \LLVM{'}s register allocator and may be lifted in the future,
when the register allocation rematerialization pass has been improved to catch
these regressions.

@@

\begin{figure}
\centering
%rankdir=LR;
\digraph[scale=0.35]{abc}{
  B0 -> B1
  B0 -> B2
  B1 -> B3
  B1 -> B4
  B2 -> B5
  B2 -> B6
}
\caption{\CFG{} to illustrate partitioning}
\label{fig:dfsin}
\end{figure}

@@

\subsection{Code generation}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed, and the \SSA{} form is restored by updating the
intermediate representation (IR) to reflect the changes. At the same time
\MemorySSA{} is also updated to get up-to-date information about memory references.

@@

\section{\SSA{} based global code motion of identical computations}
\label{sec:implementation-details}
% Move this before legality and profitability??
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna
In this section we will present an overview of the algorithm we implemented in
\LLVM{}.  The amount of hoisting depends on whether we collect \GVN{} of
instructions before finding candidates (optimistic) or, on-demand
(pessimistic). It also depends on the generality of the \GVN{} algorithm as
mentioned earlier in Section~\ref{subsec:finding-candidates}. We have
implemented a optimistic global code motion of congruent instructions which uses
the liveness analysis as illustrated in Das \cite{das2012} and ranking
expressions explained in Rosen \cite{rosen1988global}.

@@

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot not guarantee progress is marked
as a barrier. In the absense of context (as in our current implementation), some
instructions which might be safe are still classified a barrier e.g., volatile
loads/stores, calls with missing attributes.

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not compute
equivalence classes in a effective way. We found it simpler to just recompute
the value numbers and start finding hoistable candidates again.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ Graph and MergeSets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis expensive but only performed
on-demand for hoistable candidates so it does not affect the overall compile
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the code-motion, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite code-motion by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

@@

While benchmarking \SPEC{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined. Various code-size metrics are shown in
Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
instructions were sunk.
